\batcave: Calling somatic mutations with a tumor- and site-specific prior

### 2019-12-20 NAR_DRAFT.tex
commit to merge
check in changed Shi mutation rate

\subsection{Estimation of the mutation rate}
The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
\end{equation}
The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
In the R implementation of \batcave, $f_{min}$ is a free parameter.
For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
\rngcomment{Note that sample contamination by normal cells can result in underestimation of tumor mutation allele frequencies.
To account for this, our implementation of BATCAVE takes as input an estimated sample purity $p$.
All estimated allele frequencies are then multiplied by a factor of $1/p$ before being used in Eq.~\ref{eqn:mut_rate}.
}

@@

\subsection{Real data}
We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.
We downloaded the normal and primary whole-genome AML tumor bam files from dbGaP accession number phs000159.v8.p4.
Griffith et al.\ generated a platinum set of variant calls for this tumor~\cite{Griffith2015}, which we used for our true positive dataset.
We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
\rngcomment{Griffith et al.\ estimated the purity of their primary sample to be 90.7\%, and we used this value for tumor purity in \texttt{batcaver}.}
Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
For these multi-region data, we \rngcomment{used the XXX samples} and ran \batcave separately on each sequenced region and combined results to generate precision-recall curves.
\rngcomment{Shi et al.\ estimated sample purities of between XXX and XXX, so we set the purities each sample to XXX.}
We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCr37, so we used that reference.

@@

\begin{table*}[b]
  \tableparts{%
  \caption{Variant calling metrics for all data sets.}
  \label{table:01}%
  }{%
  \begin{tabular}{lllllllllll}
  \toprule
  Scenario & Mutation profile & $\mu$ & \multicolumn{2}{c}{AUROC} & \multicolumn{2}{c}{AUPRC} & \multicolumn{2}{c}{ICI} 
  \\
  & & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
  \\
  \colrule
  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & .117 & .287
  \\
  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & .118 & .214
  \\
  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & .120 & .219
  \\
  500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
  \\
  500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
  \\
  500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
  \\
  real AML \cite{Griffith2015} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
  \\
  real breast \cite{Shi2018} & Actual & 3.6e-8 & -- & -- & .972 & .972 & -- & --
  \\
  \botrule
  \end{tabular}%
  }
  {$\mu$ = per-base mutation rate, AUROC/AUPRC = Area Under Receiver Operating Characteristic / Precision-Recall Curve, ICI = Integrated Calibration Index. Smaller values of ICI are superior.}
 \end{table*}

>>>>

\subsection{Estimation of the mutation rate}
The final piece of Eq. \ref{eqn:detailed_prior} is $p(M)$, the prior probability of mutation, which we specify as the per-base per-division mutation rate $\mu$.
In an exponentially growing and neutrally evolving tumor, branching process calculations \cite{Williams2018} show that the expected total number of mutations $\mathrm{M_{tot}}$ between two allele frequencies ($f_{min}$,$f_{max}$) is
\begin{equation}
  \label{eqn:mut_rate}
  \mathrm{M_{tot}}(f_{min},f_{max}) = N\frac{\mu}{\beta}\left(\frac{1}{f_{min}} - \frac{1}{f_{max}}\right).
\end{equation}
The number of bases $N$ is $3\cdot10^9$ for a whole genome and $3\cdot10^7$ for a whole exome.
The quantity $\mu/\beta$ is the effective mutation rate, where $\beta$ is the fraction of cell divisions that lead to two surviving lineages.
We make the simplifying assumption that there is no cell death ($\beta = 1$), so we somewhat over-estimate $\mu$.
We then estimate $\mu$ by counting observed high-confidence mutations between allele frequencies $f_{min}$ and $f_{max}$.
We set $f_{max}$ to be the largest allele frequency in $\mathrm{D}$, but we must choose $f_{min}$ conservatively, depending on sequencing depth.
In the R implementation of \batcave, $f_{min}$ and tumor_purity are free parameters.
For this paper, we set $f_{min} = 0.05$, because we are working at high depth.
\rngcomment{Note that sample contamination by normal cells can result in underestimation of tumor mutation allele frequencies.
To account for this, our implementation of BATCAVE takes as input an estimated sample purity $p$.
All estimated allele frequencies are then multiplied by a factor of $1/p$ before being used in Eq.~\ref{eqn:mut_rate}.
}

@@

\subsection{Real data}
We analyzed two real data sets, one from an acute myeloid leukemia (AML)~\cite{Griffith2015} and one from a multi-region sequencing experiment in breast cancer~\cite{Shi2018}.
We downloaded the normal and primary whole-genome AML tumor bam files from dbGaP accession number phs000159.v8.p4.
Griffith et al.\ generated a platinum set of variant calls for this tumor~\cite{Griffith2015}, which we used for our true positive dataset.
We downloaded the normal and tumor whole-exome breast cancer bam files from NCBI Sequence Read Archive accession SRP070662.
\rngcomment{Griffith et al.\ estimated the purity of their primary sample to be 90.7\%, and we used this value for tumor purity in \texttt{batcaver}.}
Shi et al.\ generated a gold set of variant calls for each tumor region sequenced~\cite{Shi2018}, which we used for our true positive dataset.
For these multi-region data, we \rngcomment{used the XXX samples} and ran \batcave separately on each sequenced region and combined results to generate precision-recall curves.
\rngcomment{Shi et al.\ estimated sample purities of between XXX and XXX, so we set the purities each sample to XXX.}
We called variants using Mutect 1.1.7 as in our simulations, except that both these data sets were originally aligned to GRCrh37, so we used that reference.

@@

\begin{table*}[b]
  \tableparts{%
  \caption{Variant calling metrics for all data sets.}
  \label{table:01}%
  }{%
  \begin{tabular}{lllllllllll}
  \toprule
  Scenario & Mutation profile & $\mu$ & \multicolumn{2}{c}{AUROC} & \multicolumn{2}{c}{AUPRC} & \multicolumn{2}{c}{ICI} 
  \\
  & & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
  \\
  \colrule
  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & .117 & .287
  \\
  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & .118 & .214
  \\
  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & .120 & .219
  \\
  500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
  \\
  500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
  \\
  500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
  \\
  real AML \cite{Griffith2015} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
  \\
  real breast \cite{Shi2018} & Actual & 7e-7 & -- & -- & .972 & .972 & -- & --
  \\
  \botrule
  \end{tabular}%
  }
  {$\mu$ = per-base mutation rate, AUROC/AUPRC = Area Under Receiver Operating Characteristic / Precision-Recall Curve, ICI = Integrated Calibration Index. Smaller values of ICI are superior.}
 \end{table*}
